\chapter{Background and State-Of-The-Art}
\label{chp:background}



\section{Variable Software Development}
\label{sec:background_jolie_syntax}

% What are variable/configurable software systems? 
% How are they typically implemented in real-world?
% Do Compuer Scientist have a good understanding about them?
Moderns software systems are rarely developed as a single, fixed program with one
well defined behavior.  
Instead, many systems are designed as configurable systems in order to support
different platforms, environments and user or customer requirements within a
single code base.  
Rather than maintaining multiple separate implementation for each possible
variant, developers introduce variability into the software itself, allowing
different configurations to be derived from the same source code.

This approach is widely used in the real world software development.  
Operating systems, system libraries, embedded software and server application
often need to adapt to different hardware architecture, operating system,
performance requirements or optional features.  
For example, an operating system kernel may include different code depending on
whether it is built for a specific processor architecture, while a network server
may enable or disable certain features based on configuration options.  
Supporting such variability within a single code base makes it possible to reuse
large parts of the implementation and reduces the need to maintain separate forks
of the same system.

At the same time, configurability introduces additional complexity.  
Reasoning about a single program, developers must consider a potentially large
number of software variants.  
Each configuration represents a specific combination of features and may include
or exclude different parts of the code.  
In large systems, the number of valid configurations can easily grow into the
thousands or even millions, making it infeasible to test or inspect each variant
individually.  
As a result, developers typically reason about variability symbolically rather
than concretely, relying on conditions that describe when certain code should be
active.

In practice, variability is often implemented using conditional code.  
In languages such as C and C++, this is commonly done through preprocessor
directives such as \texttt{\#if}, \texttt{\#ifdef}, and \texttt{\#ifndef}.  
These directives are used to guard code fragments with logical conditions, also
referred to as presence conditions.  
A presence condition describes when a specific piece of code becomes part of a
software variant.  
In practice, such conditions are often expressed as logical expressions and can be
combined to represent more complex configuration rules.

While conditional compilation is a powerful mechanism, it also makes the code
harder to understand and maintain.  
Small changes to a presence condition can have far reaching consequences, as they
may affect many configurations at once.  
Developers working on a specific change often focus on the configuration they are
currently interested in and may not be fully aware of how the same change impacts
other variants of the system.  
This makes configurable software particularly open to subtle and unintended
errors.

The complexity introduced by variability becomes especially apparent during
software evolution.  
As systems evolve over time, developers continuously modify, extend and refactor
code.  
In configurable systems, such edits not only change the implementation itself but
may also alter the structure of variability, that is, the relationship between
features, conditions, and code fragments.  
For example, a seemingly minor modification to a conditional expression can
unintentionally expand or restrict the set of configurations in which a piece of
code is active.

These effects are often not visible in standard version control diffs, which
typically represent changes as simple additions or deletions of lines of code.  
As a result, developers may overlook important changes to variability, leading to
configuration-specific faults that are difficult to detect and diagnose.  
Understanding how changes to presence conditions affect the overall variability
of a system is therefore crucial for maintaining the correctness and reliability
of configurable software.


\subsection{Principles of Software Product Lines}
% What is a feature?
% What is a configuration and a software product?
%

\subsection{Classification of Edits to Configurable Software}

Edits to variable software can fully classified in 8 categories~\cite{Bittner2022_edits}.

\section{Principles of Natural Language Processing}

\subsection{Basics}

% What is the idea behind NLP?
% What techniques can be used to get information of

\subsection{Topic Modeling}
